---
title: "Introduction to the convergEU package"
author: "<br>Federico M. Stefanini"
date: "`r Sys.Date()` -  rel 1.0.0 <br><br> Index:"
output:   
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{User-Guide}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---



```{r setup, include = FALSE}
require(ggplot2)
require(dplyr)
require(tidyverse)
require(convergEU)
require(eurostat)
require(purrr)
require(tibble)
require(tidyr)
require(ggplot2)
require(formattable) 
require(kableExtra)
require(caTools)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

```


<br><br><br>


# Introduction


The *convergEU* R package is a set of S3 functions and data objects
suited for the analysis of economic and social convergence of Member
States (**MS**) withing the European Union (**EU**).     

This vignette is intended to be a gentle introduction to 
the analysis of convergence performed with *convergEU* suite of functions.

Data from Eurofound (local) and Eurostat (download) are available
from within the package with little effort.
Nevertheless, data objects may also be created from scratch,
following the R syntax to create or import  datasets.

A dataset created or processed in this package is almost always
a **tibble**, therefore at least some confidence with the  **dplyr**
package
([online dplyr site](https://dplyr.tidyverse.org/)) 
is convenient, although in some cases a bit more of
the **tidyverse**
([https://www.tidyverse.org/](https://www.tidyverse.org/))
is exploited.   
For a general introduction  see  "R for Data Science"
([online R4DS](https://r4ds.had.co.nz/)) by Wickham and Grolemund.

Imported or download data may or may not be in a tidy shape,
here defined as a rectangular table with ordered years
in the first column and with two or more MS in subsequent columns.
Such a data structure may implicitly refer to males and/or
a given age-class, a fact that may or may not be evident through
further columns in the tibble.   

The most common data to process are indicators downloaded from
EU repositories, thus it is safe to remember that an indicator
may represent a better performance-situation  of a country for high values
(**highBetter**) or when it takes very low values (**lowBetter**).
When this is a critical feature to properly process data through a function,
a specific argument must be provided, that is one string 
chosen between the two possibilities
"highBetter" and "lowBetter".  

One  general advice pertains the member states included into a dataset
(tibble object) as an input to a function.
It is always the case that a function that calculates a summary over a set 
of EU countries (like the average) must receive indicators for all
those countries within the dataset in  input,
even when the interest is focused on just two or three
countries among them.



A general feature of results produced by most
functions in the *convergEU* package is that 
a list with metainformation is returned.
It has three components: *\$res, \$msg, \$err*.
The first list component, *\$res*, is the actual result, 
if computed.
The second component, *\$msg* is a message
decorating the computed result,
possibly a warning.
The third component, *\$err*, is
an error message or a list of errors
when a result is not computed.



The **R** packages exploited in this vignette are:
```{r,eval=FALSE}
require(convergEU)
require(ggplot2)
require(dplyr)
require(tidyverse)
require(eurostat)
require(purrr)
require(tibble)
require(tidyr)
require(ggplot2)
require(formattable) 
require(caTools)
```




<br> 

#  Loading and preparing data

Two types of data sources are considered here because they are
straightforward to work with in the *convergEU* package:
data produced by Eurofound, statistically available without
and active internet connection,
and Eurostat data that can be downloaded
on the fly is while an internet connection is active.



##  Locally accessible datasets: Eurofound data

Some datasets are accessible from package  *convergEU* 
using the R function *data()*, for example:
```{r,eval=FALSE}
library(convergEU)
data("emp_20_64_MS",package = "convergEU")
head(emp_20_64_MS)
```

Eurofound datasets  EWCS and  EQLS are
locally available within  *convergEU*, see:
```{r,eval=FALSE}
data(package = "convergEU")
```
The  object `dbEUF2018meta` contains a description (metadata)
of  locally  available data from within the *convergEU* package,
without any downloading from Eurofound website. 
```{r}
print(dbEUF2018meta, n=200,width=200)             
```

The raw local Eurofound database is accessed as follows:   
```{r}
require(convergEU)
data(dbEurofound)
head(dbEurofound)
```
where variable names are:
```{r}
names(dbEurofound)
```
and the time ranges in the interval:
```{r}
c(min(dbEurofound$time), max(dbEurofound$time))
```
The database is not complete in such a time range
for all considered countries.

**NOTE: within *convergeEU* package, Eurofound data are statically stored**.
Please update this package to have the most recent version of Eurofound data. 



A special role is played by the Eurofound dataset exploited during the development of
this package: *emp_20_64_MS*.
This is the  tidy version derived  from  dataset *emp_20_64*
on the employment within age class 20 to 64:
```{r,eval=FALSE}
help(emp_20_64_MS)
```
thus such dataset is always ready for the analysis
without further processing.

 

All other indicators are extracted from the Eurofound database
in the first step of an analysis:
this is the data preparation step.
It amounts to choose a time  interval, an indicator and a set of countries
(**MS**, Member States), for example:
```{r}
convergEU_glb()$EU12
```
among those available:
```{r}
names(convergEU_glb())[c(3:8)]
```

Please note that EU19 and Eurozone are synonims:
```{r}
convergEU_glb()$EU19
```


The general procedure to select an indicator is illustrated below
for the   "lifesatisf" indicator taken from the column "Code\_in\_database"
within  the object *dbEUF2018meta* that contains  the meta information:
```{r}
head(dbEUF2018meta)
myTB <- extract_indicator_EUF(
    indicator_code = "lifesatisf", #Code_in_database
    fromTime=2003,
    toTime=2016,
    gender= c("Total","Females","Males")[2],
    countries= convergEU_glb()$EU12$memberStates$codeMS
    )
  
myTB
```
which results in a dataset ready for further analyses
in the  list component named "$res";
further components are "\$msg" that possibly carries messages
for the user and "\$err" which is a string
containing an error  message, if an error occurs.














## Downloadable data: Eurostat repository


Eurostat data are downloaded on the fly
from an active internet connection.  

The heterogeneity in the structure of different indicators
normalized into the database requires some attentions.
A list of covariates   for each indicator is sometimes present
besides age and  gender thus their values must be set
to produce a tidy dataset time by countries.
 
First, raw data may be downloaded using the option
*rawDump=T*:
```{r,echo=TRUE,eval=FALSE}
ddTB1 <- download_indicator_EUS(
      indicator_code= convergEU_glb()$metaEUStat$selectorUser[1],
      fromTime = 2005,
      toTime = 2015,
      gender= c(NA,"T","F","M")[2],#c("Total","Females","Males")
      countries =  convergEU_glb()$EU28$memberStates$codeMS,
      rawDump=T )
ddTB1
```
```{r,echo=FALSE}
sourceFile1 <- system.file("extdata", package = "convergEU")
# save(ddTB1,file=file.path(sourceFile1,"ddTB1.RData"))
load(file.path(sourceFile1,"ddTB1.RData"))
ddTB1
```


which is not a tidy dataset.
Note that *unit*   and   *isced11*  are auxilary valiables
specific for this indicator and that some filtering
must be performed to obtain a tidy dataset
years by countries.
At this purpose,  the argument  *rawDump=F* indicates
that bulk data are filtered and reshaped, as shown below:
```{r,echo=TRUE,eval=FALSE}
ddTB2 <- download_indicator_EUS(
      indicator_code= convergEU_glb()$metaEUStat$selectorUser[1],
      fromTime = 2005,
      toTime = 2015,
      gender= c(NA,"T","F","M")[1],#c("Total","Females","Males")
      ageInterv = NA,
      countries =  convergEU_glb()$EU28$memberStates$codeMS,
      rawDump=F,
      uniqueIdentif = 1)

convergEU_glb()$metaEUStat$selectorUser[1]
ddTB2
```
```{r,echo=FALSE}
convergEU_glb()$metaEUStat$selectorUser[1]
sourceFile1 <- system.file("extdata", package = "convergEU")
# save(ddTB2,file=file.path(sourceFile1,"ddTB2.RData"))
load(file.path(sourceFile1,"ddTB2.RData"))
ddTB2
```
where *convergEU\_glb()\$EU28\$memberStates\$codeMS*
is a vector of strings for the considered countries,
*convergEU\_glb()\$metaEUStat\$selectorUser[1]*
contains the name of the indicator and where
*ageInterv* may take a value when an age interval has to be specified
for a given indicator. 
The result is a list with the following components:    
  
  * \$res contains the selected data as a tidy tibble (dataset);     
  * \$msg\$gender is a component describing results; \$msg\$gender states
    that gender was not selected thus it was automatically set;   
  * \$msg\$Age states that age class was automatically set;     
  * \$msg\$Further\_Conditioning contains further variables to set in order
    to obtain a tidy dataset; in particular 
    \$msg\$Further\_Conditioning\$current states the value of conditioning
    specific  variables taken from the list of possible ones
    \$msg\$Further\_Conditioning\$available\_seleTagLs;    
  * \$msg\$Conditioning contains common conditiongn variables, such as ageInterv
    and gender.

It is therefore possible to call several times the same function and
specify the argument *uniqueIdentif* as an integer among those in the first column
left of *\$msg\$Further_Conditioning\$available\_seleTagLs*  to obtain the same
indicator under differt scales and contexts.
For example the fifth conditioning context is for males in age interval 
"Y15-64" is:
```{r,echo=T,eval=FALSE}
ddTB3 <- download_indicator_EUS(
      indicator_code= convergEU_glb()$metaEUStat$selectorUser[1],
      fromTime = 2005,
      toTime = 2015,
      gender= "M",
      ageInterv = "Y15-64",
      countries =  convergEU_glb()$EU28$memberStates$codeMS,
      rawDump=F,
      uniqueIdentif = 5)
ddTB3
```
```{r,echo=FALSE}
sourceFile1 <- system.file("extdata", package = "convergEU")
# save(ddTB3,file=file.path(sourceFile1,"ddTB3.RData"))
load(file.path(sourceFile1,"ddTB3.RData"))
ddTB3
```







<br> 

# Data preparation: from data structure to  imputation of missing values


The analysis of convergence is performed on clean and imputed data:
a tidy dataset years by countries.
The first step after downloading data is the description of the main
features of such a dataset.  

An illustrative example follows with the indicator *"JQIintensity_i"*:
```{r,eval=T}
# print(dbEUF2018meta[11,],n=20,width=100)
t(dbEUF2018meta[11,])
```

First the raw dataset is downloaded:
```{r,eval=FALSE}
ddTB4 <- extract_indicator_EUF(
    indicator_code = "JQIintensity_i", #Code_in_database
    fromTime= 1965,
    toTime=2016,
    gender= c("Total","Females","Males")[1],
    countries= convergEU_glb()$EU28$memberStates$codeMS
    )
print(ddTB4$res,n=35,width=250)
```
```{r,echo=FALSE}
sourceFile1 <- system.file("extdata", package = "convergEU")
# save(ddTB4,file=file.path(sourceFile1,"ddTB4.RData"))
load(file.path(sourceFile1,"ddTB4.RData"))
print(ddTB4$res,n=35,width=250)
```
where error messages are not shown (\$err list component empty).  



The inspection of the print output  reveals that
missing values are present.
More features may be investigated as usual with common R functions,
like dimension and variable names, but first it is convenient to
assign a meaningfull name to the downloaded data:
```{r}
JQIinte <- ddTB4$res 
dim(JQIinte)
```
that is   $5$ rows and $30$ columns, and variable names:
```{r}
names(JQIinte)
```


A dataset can't have qualitative variables, neither vector of strings
nor missing values for computing convergence measures.
A time variable should also be present, and if the name is not "time",
than it must be passed during  function calls as an argument to have
proper data processing.
The *check\_data()* function may called to check for the presence of unsuited
features that must be solved before starting the analysis.
The object returned states if the dataset is ready for calculations, and
if it is not, the error component states why checking failed:    

  * "Error: one or more missing values in the dataframe."    
  * "Error: qualitative variables  in the dataframe."    
  * "Error: string  variables in the dataframe."    
  * "Error: timeName variable absent."    
  * "Error: time variable is not ordered."    

For example, with the *JQIinte* data we have:
```{r}
check_data(JQIinte,timeName="time")
```
missing values are present thus missing imputation is required
by using  the *impute_dataset* function:
```{r}
JQIinteImp <- impute_dataset(JQIinte, timeName = "time",
                          countries=convergEU_glb()$EU28$memberStates$codeMS,
                          tailMiss = c("cut", "constant")[2],
                          headMiss = c("cut", "constant")[2])$res 
print(JQIinteImp,n=35,width=250)
```
where *\$res* was added to the function call in order 
to extract just the list component of interest.
The imputation selected for the first (tail) and last (head) years
is "constant", thus the first not missing value is propagated to missing
years, but the alternative of cutting all years in which one or more
missing are presents may be selected with the arguments: 
```
tailMiss = "cut",headMiss = "cut"
```

In case more information is needed on years where
missing values are located for a given country, say *HR*,
the following simple code helps:
```{r}
select(filter(JQIinte, is.na(HR)),time,HR) 
```
or by using pipes, after *magrittr* package:
```{r}
JQIinte %>% 
  filter(is.na(HR)) %>%
  select(time,HR)
```




The *check_data* function is called again but on imputed data:
```{r}
check_data(JQIinteImp)
```
where the suspected string variable is *sex*:
```{r}
JQIinteFin <- dplyr::select(JQIinteImp,-sex)
check_data(JQIinteFin)
```
thus  *JQIinteFin* is the final object to start the data analysis.


The *tidyverse* functions *mutate, select, filter* are the workhorses
for more elaborated selection and inspections, with or without the use of 
the new  forward-pipe operator (%>%).


Note that before starting the analysis, the **number of digits
maye be selected**, for example, if rounding to integer is preferred,
the above tibble must be changed by invoking the
function *round()* where *digits = 0*:
```{r}
JQIinteFin[, -1] <- round(select(JQIinteFin,- time), digits = 0)
JQIinteFin
```






## Imputing missing values using a straight line

The basic imputation  method is deterministic, like in the average of two
interval endpoints separated by just one year missing.
If several missing values are present in a  row 
a linear change of an indicator  is assumed over time between the two 
observed time points flanking a chunk of missing values.

```{r} 
intervalTime <-  c(1999,2000,2001,2002,2003) 
intervalMeasure <- c( 66.5, NA,NA,NA,87.2) 
currentData <- tibble(time= intervalTime, veval= intervalMeasure) 
currentData
resImputed <- impute_dataset(currentData,
                           countries = "veval",
                           timeName = "time",
                           tailMiss = c("cut", "constant")[2],
                           headMiss = c("cut", "constant")[2]) 
resImputed$res  
``` 
 
In the figure below, grey points are imputed
using observed values represented by solid blue points:
```{r,echo=FALSE,fig.width=4,fig.height=3} 
tmp <-  as.data.frame(currentData[ c(1,5),] )
tmp2 <- as.data.frame(resImputed$res[2:4,] )

myg <- ggplot(as.data.frame(resImputed$res),  mapping=aes(x=time,y=veval)) + 
  geom_point() + 
  geom_line(data=resImputed$res,col="red") + 
  geom_point(data=tmp,mapping=aes(x=time,y=veval), 
              size=4, 
              colour="blue")  + 
  geom_point(data= tmp2, 
             aes(x=time,y=veval),size=4,alpha=1/3,col="black") + 
  xlab("Time") + ylab("Measure / Index") +  
  ggtitle( "Blue points are observed values (grey ones are missing) \n") 
   
myg 

``` 
 
 
 
It must be emphasized that typical EU indicators
run along few years thus $10\%$ of missing values within a country may
already represent a context that requires substantive
reasoning before interpreting results after imputation.
The user may or may not  go ahead with the analysis
depending on the considered context.





## Weighted average smoothing of a complete dataset


It may be of interest to assume that part of the variability 
observed in a country on a given index is **not structural**,
i.e. not due to causal determinants by to transient
fluctuations.
Furthermore, the interest here is not directed towards
prediction but on smoothing values observed in the whole
considered time interval.

In such a case a smoothing procedure remove sudden large changes 
showing a less variable time serie than the original.

Given that here short time series (panel data) are considered,
a three points weighted average is proposed.
The smoother substitutes an original raw value $y_{m,i,t}$ of country $m$
indicator $i$ at time $t$ with the weighted average
$$\check{y}_{m,i,t}  = y_{m,i,t-1} ~ (1-w)/2   +w ~y_{m,i,t} +y_{m,i,t+1} ~(1-w)/2$$
where $0< w \leq 1$. The special case $w=1$ corresponds to no smoothing.
In case of missing values an NA is returned. If the weight is outside
the intervel $(0,1]$ then a NA is returned.
The first and last values are smoothed using weights $w$ and $1-w$.

After loading data, imputation takes place and finally smoothing is performed.
Now, countries IT and DE are considered to illustrate the procedure.
First check if missing values are present:
```{r}
workTB <- dplyr::select(emp_20_64_MS, time, IT,DE)
check_data(workTB)
```
thus checking is passed, so we go with the smoothing step
after deleting the time variable:
```{r}
resSM <- smoo_dataset(select(workTB,-time), leadW = 0.149, timeTB= select(workTB,time))
resSM
```
and for a comparison:
```{r}
compaTB <- select(bind_cols(workTB, select(resSM,-time)), time,IT,IT1,DE,DE1)
compaTB
```

A graphical output shows changes for "IT", with original
index in blue and smoothed index in red:
```{r,fig.width=4,fig.height=3}
qplot(time,IT,data=compaTB) + 
  geom_line(colour="navyblue") +
  geom_line(aes(x=time,y=IT1),colour="red") +
  geom_point(aes(x=time,y=IT1),colour="red",shape=8)
```
similarly for DE
```{r,fig.width=4,fig.height=3}
qplot(time,DE,data=compaTB) + 
  geom_line(colour="navyblue") +
  geom_line(aes(x=time,y=DE1),colour="red") +
  geom_point(aes(x=time,y=DE1),colour="red",shape=8)

```

A weight equal to 1 leaves data unchanged:
```{r,fig.width=4,fig.height=3}
resSM <- smoo_dataset(select(workTB,-time), leadW = 1, timeTB= select(workTB,time))
compaTB <- select(bind_cols(workTB, select(resSM,-time)), time,IT,IT1,DE,DE1)
qplot(time,IT,data=compaTB) + 
  geom_line(colour="navyblue") +
  geom_line(aes(x=time,y=IT1),colour="red") +
  geom_point(aes(x=time,y=IT1),colour="red",shape=8)
```


**A time window larger than $3$ could be considered, but
deep thoughts  are recommended about  economic and social changes
that may happen in EU during  $5$ consecutive years.**






## Moving Average smoother 


Several alternative smoothing algorithm are available in R.
Classical *ma* smoothers are described in the *caTools* package.


The emp_20_64_MS dataset is now chosen  with
Italy the country selected to illustrate operations.

```{r}
data(emp_20_64_MS)
cuTB <- select(emp_20_64_MS,time)
cuTB <- mutate(cuTB,ITori =emp_20_64_MS$IT)
```

At the begining and end of this series values are averages
calculated on a smaller and smaller number of observations (tails):
```{r}
cuTB <- mutate(cuTB, IT_k_3= runmean(emp_20_64_MS$IT, k=3, 
        alg=c("C", "R", "fast", "exact")[4],
        endrule=c("mean", "NA", "trim", "keep", "constant", "func")[4],
        align = c("center", "left", "right")[1]))

cuTB <-  mutate(cuTB, IT_k_5= runmean(emp_20_64_MS$IT, k=5, 
        alg=c("C", "R", "fast", "exact")[4],
        endrule=c("mean", "NA", "trim", "keep", "constant", "func")[4],
        align = c("center", "left", "right")[1]))

cuTB <-  mutate(cuTB, IT_k_7= runmean(emp_20_64_MS$IT, k=7, 
        alg=c("C", "R", "fast", "exact")[4],
        endrule=c("mean", "NA", "trim", "keep", "constant", "func")[4],
        align = c("center", "left", "right")[1]))
```
where options *alg,endrule,align* in the *runmean*
function are discussed in the *caTools*
package.


The Figure below shows results for different degrees of smoothing:
original (black), k=3 (red), k=5 (blue), k=7 (orange).
```{r}
myG <- ggplot(cuTB,aes(x=time,y=ITori))+geom_line()+geom_point()+
       geom_line(aes(x=time,y=IT_k_3),colour="red")+
       geom_point(aes(x=time,y=IT_k_3),colour="red")+
       #
       geom_line(aes(x=time,y=IT_k_5),colour="blue")+
       geom_point(aes(x=time,y=IT_k_5),colour="blue")+
       #
       geom_line(aes(x=time,y=IT_k_7),colour="orange")+
       geom_point(aes(x=time,y=IT_k_7),colour="orange")

myG
```


It is typically the case that 
the time serie is so short that at $k=7$ a lot of observations are smoothed with different number of observations (shorter at start and end).   


The above calulcations are performed by a function in the *convergEU* package:
```{r}
cuTB <-  emp_20_64_MS[,c("time","IT","DE")]
ma_dataset(cuTB, kappa=3, timeName= "time")
```
that is a bit less flexible but it  leads to a standard
tidy dataset.










<br>



## Absolute change


Absolute change is defined as:
$$
\Delta y_{m,i,t} = y_{m,i,t} - y_{m,i,t-1}
$$
for country $m$, indicator $i$ at time $t$.

The R function *abso_change* in this package calculates  the above quantity,
for example in the *emp_20_64_MS* dataset, which is
tidy and without missing values: 

```{r}
data(emp_20_64_MS)
mySTB <- abso_change(emp_20_64_MS, 
                        time_0 = 2005, 
                        time_t = 2010,
                        all_within=TRUE,
                        timeName = "time")
names(mySTB$res)
```
thus the above equation results in:
```{r}
mySTB$res$abso_change
```
If desired, less digits may be displayed, for example after rounding:
```{r}
round(dplyr::select(mySTB$res$abso_change,AT:UK), 5)
```



The sum of absolute values
$$
\sum_{t=t_0+1}^{} | \Delta y_{m,i,t}|  
$$
is:
```{r}
round(mySTB$res$sum_abs_change,4)
```
and such sum can be divided by the number of pair of years
so that the result is an average per pair of years:
```{r}
round(mySTB$res$average_abs_change,4)
```








<br>

## Summaries and clusters of countries

An important summary is obtained  
as  unweighted average of country values.
The cluster of considered countries may be specified
and is also stored within the function generating global
static objects and tables, called
*convergEU_glb()*.
The  illustration of this function exploits 
the *emp_20_64_MS* dataset from the  *convergEU*
package.


First note that the EU area is made by the following
MS:
```{r}
convergEU_glb()$Eurozone
convergEU_glb()$EU19
```
while labels representing the 28 MS are:
```{r}
convergEU_glb()$EU28
```

The list of known MS labels is:
```{r}
names(convergEU_glb())[3:8]
```



For example, the unweighted average
in the *emp_20_64_MS* dataset is:
```{r}
average_clust(emp_20_64_MS, 
              timeName = "time",
              cluster = "EU28")$res[,c(1,30)]
```
while for EU12 is:
```{r}
average_clust(emp_20_64_MS,timeName = "time",cluster = "EU12")$res[,c(1,30)]
```



An unknown label, like "EUspirit", causes computation error:
```{r}
average_clust(emp_20_64_MS,timeName = "time",cluster = "EUspirit")
```
and similarly for a wrong time name:
```{r}
average_clust(emp_20_64_MS,timeName = "TTime",cluster = "EU19")
```



Time series can be also plotted: 
```{r,fig.width=5,fig.height=4}
wwTB <- average_clust(emp_20_64_MS,timeName = "time",cluster = "EU28")$res[,c(1,30)]
mini_EU <- min(wwTB$EU28)
maxi_EU <- max(wwTB$EU28)

qplot(time, EU28, data=wwTB,
      ylim=c(mini_EU,maxi_EU))+geom_line(colour="navy blue")+
      ylab("emp_20_64")
```













<br>

# The analysis of convergence


Several measures of convergence have been recently 
proposed by Eurofound  (Eurofound, 2018).
In this section, each each measure is introduced
and illustrated by a numerical example.






## Beta-convergence

Let's assume that we have a tidy dataset (tibble)
years by countries.
The calculations are performed according the  following linear model:
$$
\tau^{-1}(ln(y_{m,i,t+\tau})-ln(y_{m,i,t})) = \beta_0 + \beta_1 ln(y_{m,i,t}) +\epsilon_{m,i,t}
$$
where $m$ represents the member state of EU (country), $i$ refers to an indicator of interest, $t$ is the reference time and $\tau \in \{1,2,\ldots\}$
the lenght of the time window (typically $1$ or more years).  

<!--
In the simplest case, just two time values are considered, $t$ and $t+\tau$,
while in a more  general setup all observed times in set
$\{t,t+1,\ldots,t+\tau-1, t+\tau\}$ are included into regression.
<br><br> 
-->


In   the   implementation of  function *beta_conv()*
the same reference time is maintained  across different years and
the division on the left hand side by the amount of time elasped
may be skipped argument *useTau = FALSE* is specified.     



The output of *beta_conv()* is a list
in which transformed data, the point estimate of $\beta_1$
and a standard two tails test is reported (p-value and adjusted R squared).
One tail test $H_0: \beta_1 \geq 0$ against $H_1: \beta1< 0$
might be of some interest, but it is not implemented.


Below, an example on how to invoke the function:
```{r}
require(ggplot2)
require(dplyr)
require(tibble)

empBC <- beta_conv(emp_20_64_MS, 
                 time_0 = 2002, 
                 time_t = 2006, 
                 all_within = FALSE, 
                 timeName = "time")
empBC
```
Note that *all_within = FALSE* is the default.




A plot of transformed data and the straight line may be useful:
```{r,fig.width=6,fig.height=5}
qplot(empBC$res$workTB$indic,
      empBC$res$workTB$deltaIndic,
      xlab="log-Indicator",
      ylab="Delta-log-indicator") +
  geom_abline(intercept = as.numeric(empBC$res$summary[1,2]),
              slope = as.numeric(empBC$res$summary[2,2]),
              colour = "red") +
  geom_text(aes(label=empBC$res$workTB$countries),
            hjust=0, vjust=0,colour="blue")
```
Note that label are replicated as many times as 
the number of included  subsequent years
if *all_within=TRUE* was specified while invoking
*beta_conv()*. 









<br>

## Sigma-convergence


The key concempt in sigma-convergence is variability  with respect to the mean.
Let $Y_{m,i,t}$ be the value of indicator $i$ for member state $m$ at time $t$,
and $\overline{Y}_{A,i,t}$ the average over aggregation $A$, for example $A = EU28$,
than:    

  * the average is  $\overline{Y}_{A,i,t} = n(A)^{-1}\sum_{m \in A} Y_{m,i,t}$,
    where $n(A)$ is the number of member states within aggregation $A$;       
  * the standard deviation is $s_{A,i,t} = \sqrt(n(A)^{-1} \sum_{m\in A} (Y_{m,i,t} - \overline{Y}_{A,i,t})^2)$;     
  * the coefficient of variation is $CV(A,i,t) = 100\cdot \frac{s_{A,i,t}}{\overline{Y}_{A,i,t}}$.     

For each year, the above summaries are calculated to quantify if a reduction in heterogeneity took place.



We assume that all member states in the cluster
contributing to the unweighted mean are contained into the dataset,
for example:
```{r}
mySTB <- sigma_conv(emp_20_64_MS,timeName="time")
mySTB
```

It is possible to select a time window and let the
name of time variable to the default for this dataset:
```{r}
sigma_conv(emp_20_64_MS,time_0 = 2002,time_t = 2004)
```



As a first step, the departure from the mean 
can be characterized:   
```{r}
res <- departure_mean(oriTB = emp_20_64_MS, sigmaTB = mySTB$res)
names(res$res)
res$res$departures
```
where $-1,0,1$ indicates values respectively below $-1$,
within the interval $(-1,1)$ and above $+1$.
Details  on the contribution of each MS to the variance 
at a given time $t$ is evaluate by the square of the difference
$(Y_{m,i,t} - \overline{Y}_{EU28,i,t})^2$ 
bewteen the indicator $i$ of country $m$ at time $t$
and the unweighted average over member states, say EU28:
```{r}
res$res$squaredContrib
```


It is also possible to decompose the numerator of the variance, called deviance,
at each time in order to appreciate the  percentage of 
contribution provided by each member state to the  total deviance,
$$100 \cdot \frac{(Y_{m,i,t} - \overline{Y}_{EU28,i,t})^2}{
 \sum_{m}  (Y_{m,i,t} - \overline{Y}_{EU28,i,t})^2
}$$ 
for the indicator $i$ of country $m$ at time $t$.

```{r}
res$res$devianceContrib
```
thus each row adds to $100$.


It is possible  to produce a graphical output about the
main features of country time series, as shown below:
```{r,eval=T,fig.width=6,fig.height=7}
myGG <- graph_departure(res$res$departures,
                timeName = "time",
                displace = 0.25,
                displaceh = 0.45,
                dimeFontNum = 4,
                myfont_scale = 1.35,
                x_angle = 45,
                color_rect = c("-1"='red1', "0"='gray80',"1"='lightskyblue1'),
                axis_name_y = "Countries",
                axis_name_x = "Time",
                alpha_color = 0.9
                )
myGG
```


Any selection of countries is feasible:
```{r,eval=T,fig.width=6,fig.height=7}
#myWW1<- warnings()
myGG <- graph_departure(res$res$departures[,1:10],
                timeName = "time",
                displace = 0.25,
                displaceh = 0.45,
                dimeFontNum = 4,
                myfont_scale = 1.35,
                x_angle = 45,
                color_rect = c("-1"='red1', "0"='gray80',"1"='lightskyblue1'),
                axis_name_y = "Countries",
                axis_name_x = "Time",
                alpha_color = 0.29
                )

myGG
```

 

 
 
 
## Gamma-convergence

We now introduce gamma convergence by an index
based on ranks.


Let $y_{m,i,t}$ be the value of indicator $i$ for member state $m$ at time
$t=0,1,\ldots, T$,
and $\{ \tilde{y}_{m,i,t}: m \in A )$ the ranks for indicator $i$
over member states in the  reference set $A$, for example $A = EU28$,
at a given time $t$.
The sum of ranks within member state $m$ is:
$$
 \tilde{y}^{(s)}_{m,i} = \sum_{t=0}^T  \tilde{y}_{m,i,t}
$$
thus the variance of the sum of ranks over the given interval 
$$
Var\left[ \{\tilde{y}^{(s)}_{m,i}: m \in A \} \right]
$$
may be compared to the
variance of ranks in the reference time $t=0$:
$$
Var\left[ \{\tilde{y}_{m,i,0}: m \in A \} \right]
$$


The Kendall index KI, with respect to cluster $A$ 
of member states for the indicator  $i$ over a
given time interval is:
$$
KI(A,i,T) =  \frac{Var\left[ \{\tilde{y}^{(s)}_{m,i}: m \in A \} \right]
      }{
      (T+1)^2  ~~Var\left[\{\tilde{y}_{m,i,0}: m \in A \}\right] }
$$



The measure of  gamma-convergence is obtained with ethe following function:
```{r}
gamma_conv(emp_20_64_MS,last=2016,ref=2002,timeName="time")
```



  











## Delta-convergence



Let $y_{m,i,t}$ be the value of indicator $i$ for member state $m$ at time $t$,
and $y^{(M)}_{i,t}$ the maximum value  over member states in the 
reference set $A$, for example $A = EU28$:
$$
y^{(M)}_{i,t} = max(\{ y_{m,i,t}: m \in A\})
$$

The distance of a member state $m$ from the top performer at time $i$ is:
$$
y^{(M)}_{i,t} - y_{m,i,t}
$$
thus the overall distance at time $t$, called delta, is the sum of distances over the 
reference set  $A$ of MS:
$$
\delta_{i,t} = \sum_{m \in A} (y^{(M)}_{i,t} -  y_{m,i,t})
$$
for the considered indicator $i$.


Delta-convergence can be calculated as follows:
```{r}

delta_conv(emp_20_64_MS,"time")
```





 




 

<br>

# Automated production of scoreboard and fiches


The *convergEU* package offer the possibility of
producing scoreboards and fiches in an automated way. 

<br>

## Scoreboards

The basis of scoreboard are raw values of  an indicator (level, $y_{m,i,t}$)
for MS $m$ at time $t$ for indicator $i$.
Differences among subsequent years (change) are as well important, namely
$$
y_{m,i,t} - y_{m,i,t-1}
$$
thus a function  to calculate these values may be exploited.

Let's consider the dataset *emp_20_64_MS*,
to calculate such quantities we do the following:
```{r}
data(emp_20_64_MS)
resTB <- scoreb_yrs(emp_20_64_MS,timeName = "time")
resTB
```
where the result is a list of three components:
the summary statistics, the numerical labels to indicate 
the interval of the partition a level  belongs to,
the interval of the partition a change  belongs to.

Numerical labels are assigned as follows (see 
DRAFT JOINT EMPLOYMENT REPORT FROM THE COMMISSION AND THE COUNCIL):    
* value $1$ if a the original level or change is   $y \leq m -1 \cdot s$;   
* value $2$ if a the original level or change is  $m -1\cdot s < y \leq m - 0.5\cdot s$;   
* value $3$ if a the original level or change is  $m - 0.5\cdot s< y \leq m +0.5\cdot s$;   
* value $4$ if a the original level or change is  $m +0.5\cdot s< y \leq m + 1\cdot s$;   
* value $5$ if a the original level or change is  $y > m +1\cdot s$.      


We note that there is the possibility of representing the above summaries as
coloured plots (TO DO) into scoreboards.


For the comparison of a country with the EU average,
the following steps are recommended, from raw data:

```{r,fig.height=4,fig.width=6}
# require(ggplot2)
# data(emp_20_64_MS)
selectedCountry <- "IT"
timeName <-  "time"
myx_angle <-  45

outSig <- sigma_conv(emp_20_64_MS, timeName = timeName,
           time_0=2002,time_t=2016)
miniY <- min(emp_20_64_MS[,- which(names(emp_20_64_MS) == timeName )])
maxiY <-  max(emp_20_64_MS[,- which(names(emp_20_64_MS) == timeName )])
estrattore<-  emp_20_64_MS[,timeName] >= 2002  &  emp_20_64_MS[,timeName] <= 2016
ttmp <- cbind(outSig$res, dplyr::select(emp_20_64_MS[estrattore,], -contains(timeName)))

myG2 <- 
  ggplot(ttmp) + ggtitle(
  paste("EU average (black, solid) and country",selectedCountry ," (red, dotted)") )+
  geom_line(aes(x=ttmp[,timeName], y =ttmp[,"mean"]),colour="black") +
  geom_point(aes(x=ttmp[,timeName],y =ttmp[,"mean"]),colour="black") +
#        geom_line()+geom_point()+
    ylim(c(miniY,maxiY)) + xlab("Year") +ylab("Indicator") +
  theme(legend.position = "none")+
  # add countries
  geom_line( aes(x=ttmp[,timeName], y = ttmp[,"IT"],colour="red"),linetype="dotted") + 
  geom_point( aes(x=ttmp[,timeName], y = ttmp[,"IT"],colour="red")) +
  ggplot2::scale_x_continuous(breaks = ttmp[,timeName],
                     labels = ttmp[,timeName]) +
   ggplot2::theme(
         axis.text.x=ggplot2::element_text(
         #size = ggplot2::rel(myfont_scale ),
         angle = myx_angle 
         #vjust = 1,
         #hjust=1
         ))
  
myG2
```




It is also possible to graphically show departures  in terms of  the above 
defined partition:
```{r,fig.height=6,fig.width=6}
obe_lvl <- scoreb_yrs(emp_20_64_MS,timeName = timeName)$res$sco_level_num
# select subset of time
estrattore <- obe_lvl[,timeName] >= 2009 & obe_lvl[,timeName] <= 2016  
scobelvl <- obe_lvl[estrattore,]

my_MSstd <- ms_dynam( scobelvl,
                timeName = "time",
                displace = 0.25,
                displaceh = 0.45,
                dimeFontNum = 3,
                myfont_scale = 1.35,
                x_angle = 45,
                axis_name_y = "Countries",
                axis_name_x = "Time",
                alpha_color = 0.9
                )   

my_MSstd
```











<br> 

## Country fiches


The **counvergEU** package provides a function that
automatically prepares one or more country fiches.
This function is able to create a directory along an existing path
and to copy the rmarkdown file representing the template within it.
The rmarkdown file is parameterized so that passing different parameters
the compilation takes place with different data, say different 
indicators and countries.

It is very important to prepare complete data in a tibble (dataset)
made by a time variable and as many other variables as countries
that enter into the calculation of the time average.
Failing to satisfy this requisite causes the use of a wrong mean value
at each year.
Nevertheless one key country is specified and some other countries of 
interest may be listed to better decorate graphs and compare performances.

Below, a call to the  function *go_ms_fish()* illustrates the syntax:  

```
go_ms_fish(
    workDF ='myTB',
    countryRef ='DE',
    otherCountries = "c('IT','UK','FR')",
    time_0 = 2002,
    time_t = 2016,
    tName = 'time',
    indiType = "highBest",
    aggregation= 'EU28',
    x_angle=  45,
    dataNow=  Sys.time(),
    author = 'A.Student',
    outFile = 'Germany-up2-2016', 
    outDir = "/media/fred/STORE/PRJ/2018-TENDER-EU/STEP-1/bitbucketed/tt-fish",
    indiName= 'emp_20_64_MS'
)
  
```
but it is very important to emphasize some constraints
and unusual ways to pass parameters to such a function.
In fact, note that the first argument is the working dataset
which is passed not as an R object but as a string, the name
of the dataset that must be available in the R workspace
before invoking *go_ms_fish*.    
The second argument *countryRef* is a string with the short name of
a member country that will be shown in one-country plots.
Less obvious, argument *indiType = "lowBest"* specifies if
the considered indicator is built so that a low value is good for
a country or if a high value is good (*indiType = "highBest"*).  

Of particular importance the argument *outFile* that 
can be a string indicating the name of the output file.
Similarly *outDir* is the path (unit and folders) in which the
final compiled html will be  stored. 
The sintax of the path depend on the operating system;
for example *outDir='F:/analysis/IT2018'*  indicates that
in the usb disk called 'F', within the folder 'analysis'
is located folder 'IT2018' where R will write the country fiche.
Note that a disk called 'F' must exist and also folder 'analysis'
must exist in such unit, while on the contrary folder 'IT2018'
is created by the function if it does not already exist.

Within the above mentioned output directory, besides the compiled html,
it is also stored a file called like specified by  *outFile* 
but with added the string '-workspace.RData'
that contains data and plots produced 
during the compilation of the country fiche for further
subsequent use in other technical reports.






<br>

## Indicator fiches


An auxiliary function *go_indica_fish()*
is provided in the  R package  *convergEU*
to produce an indicator fiches, where the output is an html file.
At this purpose, an output directory must be also specifed.
Note that some arguments are passed as strings instead of objects,
as described in the last section above.   


An example of syntax to invoke the procedure is:  

```
go_indica_fish(
    time_0 = 2005,
    time_t = 2010,
    timeName = 'time',
    workingDF = 'emp_20_64_MS' ,
    indicaT = 'emp_20_64',
    indiType = c('highBest','lowBest')[1],
    seleMeasure = 'all',
    seleAggre = 'EU28',
    x_angle =  45,
    data_res_download =  FALSE,
    auth = 'A.Student',
    dataNow =  '2019/05/16',
    outFile = "test_IT-emp_20_64_MS",
    outDir = "/media/fred/STORE/PRJ/2018-TENDER-EU/STEP-1/bitbucketed/tt-fish"
  )
```








<br><br>  

# References

Below the  main references are listed:   

  * Eurofound (2017), Towards a conceptual framework to monitor convergence in the European Union - evidence from desk research and expert workshop; by Massimiliano Mascherini, Martina Bisello, Hans Dubois, Franz Eiffe (Eurofound) in cooperation with ICF Consulting Services and Irene Rioboo Leston (URJC, Spain).      
  * Eurofound (2018), Upward convergence in the EU: Concepts, measurements and indicators, Publications Office of the European Union, Luxembourg; by: Massimiliano Mascherini, Martina Bisello, Hans Dubois and Franz Eiffe.    


<br><br><br><br><br><br>






